---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# mpmsim

<!-- badges: start -->
<!-- badges: end -->

`mpmsim` is a tool for generating random or semi-random matrix population models given a particular archetype, based on life history.

## Installation

You can install the development version of `mpmsim` like this:

```{r, eval = FALSE}
# install package 'remotes' if necessary
# will already be installed if 'devtools' is installed
install.packages("remotes")

# argument 'build_opts = NULL' only needed if you want to build vignettes
remotes::install_github("jonesor/mpmsim", build_opts = NULL)
```

## Usage

First remember to load the package.

```{r}
library(mpmsim)
```

Generate single MPMs.

To generate a random matrix population models (MPM) with element values based on defined life history archetypes you can use `randomMPM`. This function draws survival and transition/growth probabilities from a Dirichlet distribution to ensure that the column totals (including death) are less than or equal to 1.

Fecundity is included as a single value, or as a vector with a length equal to the dimensions of the matrix. If it is a single value, it is placed as a single element on the top right of the matrix. If it is a vector of length `n_stages` then the fertility vector spans the entire top row of the matrix.

User can specify `archetype` to constrain the MPMs. For example, in archetype 2, the survival probability (column sums of the survival matrix) is constrained to increase monotonically as the individuals advance to later stages. See `?randomMPM` for details.

```{r}
randomMPM(n_stages = 3, fecundity = 20, archetype = 2)
```

Generate a set of MPMs.

Sets of MPMs generated using the above approach can be quickly generated using `generate_mpm_set`. For example, to generate 5 MPMs from archetype 1 we can use the following code. Users can specify acceptable population growth for the set of matrices by setting the lower and upper bounds using the `lower_lambda` and `upper_lambda` arguments. This in life history analyses because, theoretically, only life histories with lambda values close to 1 will persist in nature. 

```{r}
generate_mpm_set(n = 5, n_stages = 4, fecundity = 8, archetype = 1,
                 lower_lambda = 0.9, upper_lambda = 1.1)
```
## Generate a Leslie matrix

Generating a Leslie matrix (a matrix where the stages represent discrete age classes) is handled with the `make_leslie_matrix` function. In a Leslie matrix survival is always in the lower subdiagonal and the lower-right-hand corner element, while fertility is shown in the top row. Both survival and fertility thus have a length equal to the number of stages in the model.

Users can specify both survival and fertility as either a single value or a vector of values of length equal to the dimensions of the matrix model. If these arguments are single values, the value is repeated along the survival/fertility sequence. 

```{r}
make_leslie_matrix(survival = seq(0.1, 0.45, length.out = 4), 
                   fertility = c(0,0,2.4,5), n_stages = 4, split = FALSE)
```

Generating plausible sets of Leslie matrices can then be handled by repeating the `make_leslie_matrix` command in a loop.
For example, in the following code I produce 10 Leslie matrices with increasing survival with age.

```{r}
juvSurv <- runif(n = 10, min = 0.0, max = 0.1)
adultSurv <- runif(n = 10, min = 0.4, max = 0.8)
adultFert <- rpois(10, 6)

outputMPMs <- NULL
for (i in 1:10) {
  outputMPMs[[i]] <- make_leslie_matrix(
    survival = seq(juvSurv[i], adultSurv[i], length.out = 6),
    fertility = c(0, 0, rep(adultFert[i],4)), n_stages = 6, split = FALSE
  )
}

outputMPMs
```


## Contributions

All contributions are welcome. Please note that this project is released
with a [Contributor Code of
Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html).
By participating in this project you agree to abide by its terms.

There are numerous ways of contributing.

1.  You can submit bug reports, suggestions etc. by [opening an
    issue](https://github.com/jonesor/mpmsim/issues).

2.  You can copy or fork the repository, make your own code edits and
    then send us a pull request. [Here’s how to do
    that](https://jarv.is/notes/how-to-pull-request-fork-github/).

3.  You are also welcome to email me.
