---
title: "The robustness of age from stage methods"
author: "Owen Jones"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The robustness of age from stage methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Stage-classified matrix population models (Lefkovitch models), are valuable tools for studying population dynamics, especially in cases where organism demography is dependent on ontogenetic stage and populations where age determination is difficult. While analysis of these models is often adequate, there are situations where age-related information is desired, such as with a Leslie matrix or life table. Cochran and Ellner (1992) developed methods to obtain age-related life history traits by breaking down the transition matrix into birth, survival, and fission matrices. These methods allow for the calculation of the discrete survivorship and maternity functions. From these two fundamental trajectories, various life history metrics can be calculated including mean age at maturity, generation time, age-specific reproductive value, and senescence measures.

But how well do these methods really perform? For example, do they perform equally well for all functional forms of mortality? Can they capture differences in senescence rate? 

The following vignette illustrates how `mpmsim` can be used to address one of these questions.


## Age-from-stage

First we load the required packages.

```{r, message=FALSE}
library(mpmsim)
library(dplyr)
library(Rage)
```

Next we will use the `model_survival()` function create a `list` of age-specific survival trajectories based on a Gompertz mortality function with senescence rates (`b_1_values`) varying between 0.1 and 0.9. These values are the basis for our life tables.

```{r}
b_1_values <- seq(0.1, 0.9, 0.1)
lifeTables <- list()
for (i in 1:length(b_1_values)) {
  lifeTables[[i]] <- model_survival(
    params = c(b_0 = 0.1, b_1 = b_1_values[i]),
    model = "Gompertz"
  )
}
```

For this analysis we want to convert these age-based trajectories into stage-based matrix models. To do that we will divide the yearly age entries into stages. We do this in an arbitrary way by dividing the age vector into 3 parts: a quarter for juveniles, a half for adult, and another quarter for old individuals.

```{r}
for (i in 1:length(lifeTables)) {
  lifeTables[[i]] <- lifeTables[[i]] |>
    mutate(stage = ifelse(x <= round(max(x) * 0.25), 1,
      ifelse(x <= round(max(x) * 0.75), 2, 3)
    ))
}
```


Next we'll add fertility to those life tables.

```{r}
for (i in 1:length(lifeTables)) {
  lifeTables[[i]] <- lifeTables[[i]] |>
    mutate(fert = model_fertility(
      age = x, params = c(A = 3),
      maturity = min(x[stage == 2]),
      model = "step"
    ))
}
```

Let's take a look at one of these life tables so that the data preparation so far is clear. The most important columns here are age (`x`), survival (`px`), fertility (`fert`) and `stage` (see `?mpm_to_table` for an explanation of the columns in the life table).

```{r}
lifeTables[[5]]
```


The next step is to turn these life tables into matching pairs of (1) Leslie matrices and (2) stage-based matrices.

### Leslie matrices

First Now we can turn these life tables containing age-specific survival and fertility trajectories into Leslie matrices using the `make_leslie_mpm` function. These matrix models can be large or small depending on the maximum life span of the population: the life tables are, by default, constrained to extend to the age where the survivorship function falls below 0.01.

```{r}
leslie_matrices <- list()
for (i in 1:length(lifeTables)) {
  leslie_matrices[[i]] <- make_leslie_mpm(
    survival = lifeTables[[i]]$px,
    fertility = lifeTables[[i]]$fert,
    n_stages = nrow(lifeTables[[i]]), split = TRUE
  )
}
```

Let's examine the Leslie matrix that matches the life table from the previous section.

```{r}
leslie_matrices[[5]]$matA
```


### Lefkovitch matrices

From these Leslie matrices we can now construct the stage-structured matrix models (Lefkovitch models) using the `mpm_collapse()` function from `Rage`.

```{r}
collapsed_matrices <- list()
for (i in 1:length(lifeTables)) {
  stages <- lifeTables[[i]]$stage
  matrices <- leslie_matrices[[i]]
  collapse_list <- split(stages, stages)
  # get the indices of each element in the original vector
  collapse_list <- lapply(collapse_list, function(x) which(stages %in% x))

  collapsed_matrices[[i]] <- Rage::mpm_collapse(
    matU = matrices$mat_U,
    matF = matrices$mat_F, collapse = collapse_list
  )
}
```



Let's examine the Lefkovitch matrix that matches the Leslie matrix and life table from the previous sections:

```{r}
collapsed_matrices[[5]]$matA
```


## Recovering age-specific trajectories from the stage-based model

Next we can use age-from stage-approximation to see how well we can recover the "true" values in the original Leslie matrix/life table. In the plot, the red curve represents the "true" values of survivorship from the life table and the black curve represents the estimation of survivorship using age-from-stage approximation. 


```{r}
recovered_life_tables <- list()
for (i in 1:length(lifeTables)) {
  m1 <- collapsed_matrices[[i]]
  recovered_life_tables[[i]] <- Rage::mpm_to_table(
    matU = m1$matU, matF = m1$matF,
    remove_final = TRUE
  )
}
```



```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
i <- 5
recovered_lt <- recovered_life_tables[[i]]

plot(0:(length(recovered_lt$lx) - 1),
  recovered_lt$lx,
  type = "l",
  xlab = "age", ylab = "survivorship", main = "Survivorship"
)
lines(lifeTables[[i]]$x, lifeTables[[i]]$lx, type = "l", col = "red")
```

## Mortality and longevity

We could also look at other life table trajectories such as mortality (hazard).

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
plot(0:(length(recovered_lt$lx) - 1),
  recovered_lt$hx,
  type = "l",
  xlab = "age", ylab = "mortality", main = "mortality", ylim = c(0, 2)
)
lines(lifeTables[[i]]$x, lifeTables[[i]]$hx, type = "l", col = "red")
```


This plot clearly shows that mortality (hazard) is underestimated across most of the life course and that the trajectory is markedly different from the underlying Gompertz trajectory. This is caused by the known issue of the population structure reaching its stationary state: When the population reaches this state, the rate of change in the mortality/survival trajectory will asymptote to zero. 

We can now examine how the over-estimation of life span using age-from-stage methods is related to the Gompertz parameter like this.

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
df1 <- data.frame(b_1_values, lifespan_lt = NA, lifespan_afs = NA)
for (i in 1:length(lifeTables)) {
  df1$lifespan_lt[i] <- max(lifeTables[[i]]$x)
  df1$lifespan_afs[i] <- max(recovered_life_tables[[i]]$x)
}

df1 <- df1 %>%
  mutate(lifespan_diff = lifespan_afs - lifespan_lt) %>%
  mutate(lifespan_diff_perc = 100 * (lifespan_diff / lifespan_lt))

plot(df1$b_1_values, df1$lifespan_diff_perc,
  type = "b",
  ylab = "Lifespan overestimation (%)", xlab = "Gompertz parameter"
)
```

## Further questions?

As explained elsewhere, this issue can be ameliorated by examining the convergence to the quasi-stationary state (QSD) (see `?Rage::qsd_converge()`) and it affects both mortality/survival and fertility. One could use the methods outlined here to explore these issues more thoroughly by (for example) checking how age-from-stage methods perform with different life histories and different functional forms of mortality (see `?mpmsim::model_survival()`) and fertility.
(`?mpmsim::model_fertility()`).
