---
title: "Error propagation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Error propagation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mpmsim)
```

## Introduction

Uncertainty in the individual matrix elements of a matrix population model can lead to sampling errors, which propagate and affect the accuracy of estimates of MPM-derived metrics. These metrics may include population growth rate, generation time, and reactivity.

One way to estimate this uncertainty is by resampling the matrix model multiple times, based on assumptions about the underlying processes in the matrix model and the uncertainties in the individual matrix elements. For example, we can assume that the fertility component of the matrix is the mean of multiple Poisson processes, and that survival can be modeled as a binomial process.

The `compute_ci()` function allows users to estimate a 95% confidence interval for any metric typically derived from a matrix population model. This function resamples the matrix model based on the above assumptions and calculates the posterior distribution of the metric of interest. The lower and upper bounds of the 95% CI are then estimated from this distribution. The width of the CI reflects the precision of the estimate; a narrower CI indicates greater precision, while a wider CI indicates less precision.

Various factors can influence the width and shape of the CI, such as the sample size used to estimate the individual matrix elements, the structure of the matrix model, the underlying assumptions made by the `compute_ci()` function, and the distribution of uncertainties in the individual matrix elements. It is crucial to consider these factors when interpreting the results of a matrix population model and assessing the precision of the resulting estimates.

## Estimate 95% CI

Let's consider a matrix model **A** that is composed of submatrices **U** and **F**. We need a "split" **A** matrix because the undertlying statistical properties are different in reproduction and survival.

```{r}
matU <- matrix(c(
  0.1, 0.0,
  0.2, 0.4
), byrow = TRUE, nrow = 2)

matF <- matrix(c(
  0.0, 3.0,
  0.0, 0.0
), byrow = TRUE, nrow = 2)
```

If we know that the sample size used to estimate each element of the matrix was `20` individuals, we can estimate the downstream 95% confidence intervals for, say lambda, as follows:


```{r}
compute_ci(mat_U = matU, mat_F = matF, sample_size = 20, FUN = popdemo::eigs, what = "lambda")
```

Note that it is not necessary to use the same sample size for each element: the sample size information can be provided as a matrix of integers so that it can vary across the matrix. 

We can examine the posterior distribution of these lambda estimates estimates by using the argument `dist.out = TRUE`.

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
distLambda_20 <- compute_ci(mat_U = matU, mat_F = matF, 
                            sample_size = 20, FUN = popdemo::eigs, what = "lambda", 
                            dist.out = TRUE)
hist(distLambda_20$estimates)
```

## Exploring the impact of sample size

What if the sample size was considerably larger, at 100?

```{r, fig.height = 4, fig.width = 6, fig.align = "center"}
distLambda_100 <- compute_ci(mat_U = matU, mat_F = matF, 
                             sample_size = 100, FUN = popdemo::eigs, what = "lambda", 
                             dist.out = TRUE)
```

Comparing those two distributions, it is unsurprising that the posterior distribution with a large sample size is narrower than that of a small sample size.

```{r compareDist2, fig.height = 6, fig.width = 6, fig.align = "center"}
par(mfrow = c(2, 1))
hist(distLambda_20$estimates, xlim = c(0, 1.75))
hist(distLambda_100$estimates, xlim = c(0, 1.75))
```

What if the sample size was higher for some elements than others? No problem, the `sample_size` argument accepts a matrix of sample sizes (or a list of matrices labelled as `mat_U_ss` and `mat_F_ss`).

```{r}
sample_size_mat <- matrix(c(
  20, 100,
  20, 20
), byrow = TRUE, nrow = 2)

distLambda_variable <- compute_ci(mat_U = matU, mat_F = matF, 
                                  sample_size = sample_size_mat, 
                                  FUN = popdemo::eigs, what = "lambda", 
                                  dist.out = TRUE)
hist(distLambda_variable$estimates)
```

```{r compareDist3, fig.height = 9, fig.width = 6, fig.align = "center"}
par(mfrow = c(3, 1))
hist(distLambda_20$estimates, xlim = c(0, 1.75))
hist(distLambda_100$estimates, xlim = c(0, 1.75))
hist(distLambda_variable$estimates, xlim = c(0, 1.75))
```

If sample size is set to 0, the computation assumes that the estimate is known without error.

```{r}
sample_size_mat <- matrix(c(
  0, 0,
  0, 0
), byrow = TRUE, nrow = 2)

distLambda_variable <- compute_ci(mat_U = matU, mat_F = matF, 
                                  sample_size = sample_size_mat, 
                                  FUN = popdemo::eigs, what = "lambda", 
                                  dist.out = TRUE)
hist(distLambda_variable$estimates)
```
