---
title: "Generate a set of Leslie matrices"
author: "Owen Jones"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate a set of Leslie matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(42)
```

## Introduction

Leslie matrix models, named after Patrick Leslie, are a type of matrix population model used to describe demography of age structured populations. They are commonly used in studies of wildlife,conservation and evolutionary biology.

In a Leslie matrix the square matrix is used to model discrete, age-structured population growth with a projection interval (time step) typically representing years. Each element in the matrix represents a transition probability between different age classes or the fertility of the age class. The information in matrix model can be split into two submatrices, representing reproduction (the fertility vector) and survival/growth respectively.

**Fertility**: The first row of the Leslie matrix contains the reproductive rates of each age class. These entries indicate how many new individuals each age class can expect to produce in each projection interval.

**Survival Probabilities**: The subdiagonal (immediately below the main diagonal) of the Leslie matrix consists of survival probabilities. Each entry here shows the probability that an individual of one age class will survive to the next age class.

All other entries in the matrix are typically zero, indicating that those transitions are impossible.

To project the population size an structure through time, the Leslie matrix is multiplied by a vector that represents the current population structure (number of individuals in each age class). The result is a new vector showing the predicted structure of the population in the next time step. This process can be repeated to model population changes over multiple time periods.

Leslie matrices are useful for studying population dynamics under different scenarios, such as changes in survival rates, fecundity rates, or management strategies, and they have been widely applied in both theoretical and applied ecology.

## Aims

The purpose of this vignette is to illustrate how to simulate sets of Leslie MPMs based on known functional forms of mortality and fertility. There are several reasons why one would want to do this, including, but not limited to:

*Exploring how senescence parameters influence dynamics.
*To obtain MPMs based on parameter estimates of mortality and fertility obtained using other methods.
*To produce MPMs with known properties for teaching purposes.

It is often useful to generate sets of MPMs, rather than single matrices. For example, it is useful to explore how outcomes of modelling vary across a range of likely parameter values for mortality or fertility. Therefore, we will here create

This is done in four steps: (1) generate actuarial life tables based on functional forms of mortality; (2) add reproduction information to these life tables, based on functional forms of fertility; (3) create matrix population models from those life tables; (4) combine the life matrix population models into a convenient format.

Before starting, we load the required packages.

```{r, message=FALSE}
library(mpmsim)
library(dplyr)
library(Rage)
library(ggplot2)
```

## Functional forms of mortality and fertility

There are numerous published and well-used functional forms used to describe how mortality risk changes with age. The `model_mortality` function currently handles 6 of these models. These are: Gompertz, Gompertz-Makeham, Weibull, Weibull-Makeham, Siler and Exponential. 

In a nutshell:

*Gompertz: A mortality rate that increases exponentially with age.
*Gompertz-Makeham: A mortality rate that increases exponentially with age, with an additional age-independent constant mortality.
*Weibull: A mortality rate that scales with age, increasing at a rate that can either accelerate or decelerate, depending on the parameters of the model.
*Siler: A mortality model that separates mortality rates into two age-related components — juvenila mortality, which declines exponentially with age and adult mortality, which increases exponentially.
*Exponential: Constant mortality that is unchanging with age.

These are illustrated below.

```{r echo = FALSE, message=FALSE}
library(patchwork)
A <- model_mortality(
  params = c(b_0 = 0.01, b_1 = 0.5),
  model = "Gompertz"
)

plotA <- ggplot(A,aes(x = x, y = hx)) + 
  geom_line() + 
  xlab("Age") + 
  ylab("Hazard") + 
  ggtitle("A) Gompertz") 


B <- model_mortality(
  params = c(b_0 = 0.01, b_1 = 0.5, C = 0.2),
  model = "GompertzMakeham"
)

plotB <- ggplot(B,aes(x = x, y = hx)) + 
  geom_line() + 
  xlab("Age") + 
  ylab("Hazard") + 
  ggtitle("B) Gompertz-Makeham") + 
  geom_hline(yintercept = 0.2,linetype =2 ) + 
  coord_cartesian(ylim = c(0, NA))


C <- model_mortality(
  params = c(a_0 = 0.15, a_1 = 0.3, C = -0.11, b_0 = 0.1, b_1 = 0.1),
  model = "Siler"
)

plotC <- ggplot(C,aes(x = x, y = hx)) + 
  geom_line() + 
  xlab("Age") + 
  ylab("Hazard") + 
  ggtitle("C) Siler") + 
  coord_cartesian(ylim = c(0, NA))

D <- model_mortality(
  params = c(b_0 = 1.51, b_1 = 0.15),
  model = "Weibull"
)

plotD <- ggplot(D,aes(x = x, y = hx)) + 
  geom_line() + 
  xlab("Age") + 
  ylab("Hazard") + 
  ggtitle("D) Weibull") 

E <- model_mortality(
    params = c(b_0 = 1.51, b_1 = 0.15, C = 0.05),
  model = "WeibullMakeham"
)

plotE <- ggplot(E,aes(x = x, y = hx)) + 
  geom_line() + 
  xlab("Age") + 
  ylab("Hazard") + 
  ggtitle("F) Weibull-Makeham") + 
  geom_hline(yintercept = 0.05,linetype =2 ) + 
  coord_cartesian(ylim = c(0, NA))

FF <- model_mortality(
  params = c(b_0 = 0.1),
  model = "Exponential"
)

plotFF <- ggplot(FF,aes(x = x, y = hx)) + 
  geom_line() + 
  xlab("Age") + 
  ylab("Hazard") + 
  ggtitle("E) Exponential") 


plotA + plotB + plotC + plotD + plotE + plotFF

```

In addition to these fucntional forms of mortality, there are of course functional forms that habve been used to model fertility. The `model_fertility` function handles five types: logistic, step, von Bertalanffy, Normal and Hadwiger.


*Logistic: Fertility initially increases rapidly with age then slows to plateau as it approaches a maximum value.
*Step: Fertility is initially zero, then jumps to a particular level at a specified age, after which it remains constant.
*von Bertalanffy: This model is often used in growth dynamics but has been adapted for fertility to describe changes over age or time following a logistic growth form not limited by a strict upper asymptote. It shows how fertility rates might increase and then decrease, following a more smoothed, sigmoid curve.
*Normal : Fertility is modelled as normal distribution to describe how fertility rates increase, peak, and then decreases in a bell curve around a mean age of reproductive capacity.
*Hadwiger: The outcomes of this model is qualitatively similar to the normal distribution.

Collectively, these mortality and fertility functions offer a large scope for modelling the variaty of demographic trajectories apparent across the tree of life.

## Generate actuarial life tables

To generate actuarial (mortality) life tables, we will use the `model_mortality()` function create a `list` of age-specific survival trajectories based on a Gompertz mortality function with senescence rates (`b_1_values`) drawn randomly from a uniform distribution ranging between 0.1 and 0.9. These values are the basis for our life tables.

```{r}
b_1_values <- seq(0.1, 0.9, 0.1)
lifeTables <- list()

for (i in seq_along(b_1_values)) {
  lifeTables[[i]] <- model_mortality(
    params = c(b_0 = 0.1, b_1 = b_1_values[i]),
    model = "Gompertz"
  )
}
```

The `model_mortality` function here, by default, constrain the life table to extend to the age where the survivorship function falls below 0.01. This can be modified using the `truncate` argument. The final row of the life table is open, which is to say that the 

Although in this case, we use the Gompertz mortality function, there are several other functional forms that can be used. These include the Gompertz-Makeham model, constant mortality (also known as "exponential"), and Siler (also known as "bathtub", due to its distinctive shape).

## Add reproduction to the life tables

```{r}
for (i in seq_along(lifeTables)) {
  lifeTables[[i]] <- lifeTables[[i]] |>
    mutate(fert = model_fertility(
      age = x, params = c(A = 3),
      maturity = 3,
      model = "step"
    ))
}
```

Let's take a look at one of these life tables so that the data preparation so far is clear. The most important columns here are age (`x`), survival (`px`), fertility (`fert`) and `stage` (see `?mpm_to_table` for an explanation of the columns in the life table).

```{r}
lifeTables[[5]]
```


The next step is to turn these life tables into matching pairs of (1) Leslie matrices and (2) stage-based matrices.

### From life table to MPM

Now we can turn these life tables containing age-specific survival and fertility trajectories into Leslie matrices using the `make_leslie_mpm` function. These MPMs can be large or small depending on the maximum life span of the population. 


```{r}
leslie_matrices <- list()

for (i in seq_along(lifeTables)) {
  leslie_matrices[[i]] <- make_leslie_mpm(
    survival = lifeTables[[i]]$px,
    fertility = lifeTables[[i]]$fert,
    n_stages = nrow(lifeTables[[i]]), split = TRUE
  )
}
```


Let's examine the Leslie matrix that matches the life table from the previous section.

```{r}
leslie_matrices[[5]]$mat_A
```

